---
title: 'Data 621 Hw #5'
author: 'Group #3'
date: "2024-04-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Wine Data

### Packages

```{r}
library(dplyr)
library(summarytools)
library(corrplot)
```



### Importing Data

```{r}
wine_train_raw=read.csv(url("https://raw.githubusercontent.com/sleepysloth12/data621_hw5/main/wine-training-data.csv"))

wine_test_raw=read.csv(url("https://raw.githubusercontent.com/sleepysloth12/data621_hw5/main/wine-evaluation-data.csv"))


```



## Part I


### Summary Statistics and Distribution

```{r}
summary1 = dfSummary(wine_train_raw[, -which(names(wine_train_raw) == "INDEX")])


print(summary1, method = "render")

view(summary1)
```


### Correlation Matrix

```{r}
numeric_data = wine_train_raw %>% 
  select(-TARGET) %>% 
  select_if(is.numeric)  

correlation_matrix = cor(numeric_data, use = "complete.obs")  

print(correlation_matrix)



corrplot(correlation_matrix, method = "color", type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black")
```

Weak Correlations Mostly: 

Most variables exhibit very weak correlations with each other, which are close to 0. This suggests no strong linear relationships between most pairs of features.

LabelAppeal and STARS: The strongest correlation observed is between LabelAppeal and STARS (0.3189). This suggests a moderate positive relationship, where wines with more appealing labels tend to be rated higher by experts.


FixedAcidity and AcidIndex: There's a noticeable positive correlation between FixedAcidity and AcidIndex (0.1542). Although not very strong, it indicates that as fixed acidity increases, the overall acidity index tends to increase as well.


Alcohol and STARS: Alcohol content shows a modest correlation with STARS (0.0649), suggesting that higher alcohol content could be slightly associated with higher ratings by experts.


Negative Correlations: There are several mild negative correlations, such as AcidIndex and STARS (-0.0955), indicating that higher acid index values may slightly correlate with lower expert ratings.

## Part II

### Looking For Missing Values and Fixing it 

The following are missing:

616 Records of `ResidualSugar` (4.8%)
638 Records of `Chlorides` (5%)
647 Records of `FreeSulfurDioxide` (5.1%)
682 Records of `TotalSulfurDioxide` (5.3%)
395 Records of `pH` (3.1%)
1210 Records of `Sulphates` (9.5%)
653 Records of `Alcohol` (5.1%)
3359 Records of `STARS` (26.3%)

All of these columns (with the exception of `STARS`) are normally distributed. 

To fill in the missing values, I will use mean and standard deviation by using the `rnorm()` function and creating a function that will process these columns to fill in the missing data.

```{r}
fill_missing_with_normal = function(data, column_name) {
 
  column_data = data[[column_name]]
  
  mean_val = mean(column_data, na.rm = TRUE)
  sd_val = sd(column_data, na.rm = TRUE)
  
  num_missing = sum(is.na(column_data))
  
  new_values = rnorm(num_missing, mean = mean_val, sd = sd_val)
  
  data[[column_name]][is.na(column_data)] = new_values
  
  return(data)
}

columns_with_missing = c("ResidualSugar", "Chlorides", "FreeSulfurDioxide",
                          "TotalSulfurDioxide", "pH", "Sulphates", "Alcohol")

for (column in columns_with_missing) {
  wine_train_raw = fill_missing_with_normal(wine_train_raw, column)
}


summary2 = dfSummary(wine_train_raw[, -which(names(wine_train_raw) == "INDEX")])


print(summary2, method = "render")

view(summary2)
```

Now, there are no more missing values for the normally distributed columns.

Only 3359 (26.3%) of records now are missing `STARS` label.

I will use the existing probabilities of the distribution of `STARS` data to do bootstraping to generate the missing values.

```{r}

fill_missing_categorical = function(data, column_name, levels, probabilities) {
  
  column_data = data[[column_name]]
  
 
  num_missing = sum(is.na(column_data))
  
  
  new_values = sample(levels, num_missing, replace = TRUE, prob = probabilities)
  
  
  data[[column_name]][is.na(column_data)] <- new_values
  
  return(data)
}


stars_levels = c(1, 2, 3, 4)


stars_probabilities = c(32.2, 37.8, 23.4, 6.5) / 100


wine_train_raw = fill_missing_categorical(wine_train_raw, "STARS", stars_levels, stars_probabilities)



summary3 = dfSummary(wine_train_raw[, -which(names(wine_train_raw) == "INDEX")])


print(summary3, method = "render")

view(summary3)

```

No missing values, and we keep the distribution patterns that originally existed. Nice!


### Transforming the Data

Adding interaction terms can help to capture the combined effects of variables.

Acidity Interactions: We will explore interactions between `FixedAcidity`, `VolatileAcidity`, and `CitricAcid`.

`Sulfur Dioxide` and `Free Sulfur Dioxide`: These could interact with other chemical properties like `pH` and `Alcohol` content, affecting wine stability and taste.

Sulfur Ratio: The ratio of `FreeSulfurDioxide` to `TotalSulfurDioxide` might give insights into how bound versus free sulfur dioxide impacts wine quality.

Sugar to Acid Ratio: We will use `ResidualSugar ` to `TotalAcidity` ( `TotalAcidity` as a sum of `FixedAcidity` and other acid measures).

```{r}

wine_train_clean = wine_train_raw %>%
  mutate(
    
    FixedAcid_VolatileAcid = FixedAcidity * VolatileAcidity,
    Alcohol_CitricAcid = Alcohol * CitricAcid,
    Alcohol_pH = Alcohol * pH,
    
    
    SugarToAcidRatio = ResidualSugar / (FixedAcidity + CitricAcid), 
    FreeToTotalSulfur = FreeSulfurDioxide / TotalSulfurDioxide,
    
    
    Acid_Alcohol_Index = (FixedAcidity * VolatileAcidity) / Alcohol
  )


wine_train_clean = wine_train_clean %>%
  select(TARGET,  
         FixedAcid_VolatileAcid,  
         Alcohol_CitricAcid,     
         Alcohol_pH,             
         SugarToAcidRatio,        
         FreeToTotalSulfur,       
         Acid_Alcohol_Index,      
         AcidIndex, Alcohol, Chlorides, CitricAcid, Density,
         FixedAcidity, FreeSulfurDioxide, LabelAppeal, pH,
         ResidualSugar, STARS, Sulphates, TotalSulfurDioxide, VolatileAcidity
        )


```

### New Correlation Matrix

```{r}
cor_matrix2 = cor(wine_train_clean %>% select(-TARGET), use = "complete.obs")



```

Given the high correlations between the interaction terms and their original variables, I will remove original variables.

```{r}
wine_train_clean = wine_train_clean %>%
  select(TARGET,  
         FixedAcid_VolatileAcid,  
         Alcohol_CitricAcid,     
         Alcohol_pH,             
         SugarToAcidRatio,        
         FreeToTotalSulfur,       
         Acid_Alcohol_Index,      
         AcidIndex,  Chlorides, Density,
          FreeSulfurDioxide, LabelAppeal,
         ResidualSugar, STARS
        )
```

